\pdftitle

### 2018-05-14 maxNt_v2.tex
V2

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-14T14:56:53+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

Maximum-entropy probability models
\cite{jaynes1957,jaynes1963,sivia1996_r2006,meadetal1984} have been used in
neuroscience with this kind of purpose, for recordings from brain areas as
diverse as retina and motor cortex, at least since the 1990s
\cite{mackay1991,martignonetal1995,bohteetal2000,amarietal2003,schneidmanetal2006,shlensetal2006,mackeetal2009b,roudietal2009c,tkaciketal2009,gerwinnetal2009,mackeetal2011,mackeetal2011b,ganmoretal2011,granotatedgietal2013,tkaciketal2014b,moraetal2015,shimazakietal2015}.
Part of this literature focuses on the range of applicability of \me\
distributions
\cite{tkaciketal2006,roudietal2009,roudietal2009b,barreiroetal2010b,mackeetal2013,rostamietal2016_r2017}.
We assume that our readers are at least vaguely familiar with this kind of model.
In brief, it builds a probability distribution for neuronal activity using
only data about simultaneous firings of two neurons, or of three, and so
on. Let us take the case of the first two moments as an example. The
maximum-entropy distribution can be derived in three main ways:
\begin{enumerate}
\item From the maximum-entropy method
  \cite{jaynes1957,jaynes1963,sivia1996_r2006}. The distribution is chosen
  as the one with the highest Shannon entropy among those having particular
  first and second moments. This distribution is deemed to be
  \enquote{maximally noncommittal} \cite{jaynes1963}, but this adjective
  means little without further technical characterization. Two
  characterizations are given by the next two derivations.

\item As the asymptotic distribution of a Bayesian model called
  \enquote{exchangeable}, with a particular prior
  \cites{jaynes1986d_r1996,skilling1989b,skilling1990,rodriguez1989,rodriguez2002,portamana2017}[see
  also:][\chap~VI]{schroedinger1946_r1948}%[pp.~52--55]
  {jaynes1979b}. This distribution is the limit, when the number of time
  bins is very large, of
  \begin{equation}
    \label{eq:predictive_exchangeability}
    \p(\text{\small activity in one time bin} \|
    \text{\small 1st \amp\ 2nd empirical moments from other bins}%,\text{\small other info}
    ),
  \end{equation}
  calculated with Bayes's theorem from a particular initial joint
  distribution in which different time bins are almost independent. This
  derivation produces a probability just by taking into account those two
  moments, but does not exclude that other information might be relevant.
  It can therefore be compared with a probability that uses additional
  information, or with empirical frequencies.
\item As the asymptotic distribution of a Bayesian model based on
  sufficient statistics \cite{jaynes1982,%[\sect~I]
    jaynes1982b,portamana2017}. This distribution is the limit, when the
  number of time bins is very large, of
\begin{equation}
  \label{eq:general_prediction}
  \p(\text{\small activity in one time bin} \|
  \text{\small activities in other bins}%,\text{\small other info}
  )
\end{equation}
under the assumption that \emph{only} the values of the first and second
empirical moments of the activities are relevant for the probability above,
\ie\
\begin{multline}
  \label{eq:informal_sufficiency}
  \p(\text{\small activity in one time bin} \|
  \text{\small activities in other bins}%,\text{\small other info}
  ) \equiv{}\\
  \p(\text{\small activity in one time bin} \|
  \text{\small 1st \amp\ 2nd empirical moments from other bins}%,\text{\small other info}
  );
\end{multline}
in other words, the assumption that the first and second moments are
sufficient statistics. This limit generally does not depend on the initial
joint distribution. This derivation produces a probability by explicitly
assuming that only the first and second moments are informationally
relevant: any other information would lead to the same conditional
probability. It can therefore only be compared with empirical frequencies.
\end{enumerate}
We refer our readers to the cited references for a deeper discussion of
these derivations. It is remarkable that all three lead to the same
distribution. We shall now see, however, that if we apply these derivations
more carefully they lead to different distributions.

@@

This kind of investigation of recorded samples of neuronal activity would
have little significance if we did not expect its results to generalize
beyond the particular recorded sample. In other words, we implicitly or
explicitly expect that the conclusions drawn with \me\ methods about the
recorded neurons should somehow generalize or extrapolate at least to whole
areas those neurons are recorded from, \eg\ retina or motor cortex. We are
thus assuming that the recorded sample is a \emph{representative sample} of
that area. Whether such extrapolation makes biological sense is a difficult
question, to be analysed case by case. For the moment let's assume that it
does make biological sense, and that our sample can be assumed to be
representative of its brain area.

Then a problem appears in the way the \me\ method is typically applied, in
the literature, to recorded or artificial samples. In this note we want to
discuss this problem and offer a solution to it.

The problem is this. We can apply the \me\ method to the sample, choosing
some sufficient statistics, and generate a probability distribution for the
state of the sample. But we consider our sample to be representative of a
larger population, in which it is biologically and functionally embedded.
We can apply the \me\ method to the larger population, using the same kind
of sufficient statistics, and generate a probability distribution for its
larger state, and then find the distribution for the sample by
marginalization. Either application seems to have some commendable
features. However, \emph{the distributions obtained from these two
  applications differ}. It goes without saying that if one is
\enquote{noncommittal}, the other must be \enquote{committal}. Which choice
is most meaningful?

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-14T15:43:37+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

Maximum-entropy probability models
\cite{jaynes1957,jaynes1963,sivia1996_r2006,meadetal1984} have been used in
neuroscience with this kind of purpose, for recordings from brain areas as
diverse as retina and motor cortex, at least since the 1990s
\cite{mackay1991,martignonetal1995,bohteetal2000,amarietal2003,schneidmanetal2006,shlensetal2006,mackeetal2009b,roudietal2009c,tkaciketal2009,gerwinnetal2009,mackeetal2011,mackeetal2011b,ganmoretal2011,granotatedgietal2013,tkaciketal2014b,moraetal2015,shimazakietal2015}.
Part of this literature focuses on the range of applicability of \me\
distributions
\cite{tkaciketal2006,roudietal2009,roudietal2009b,barreiroetal2010b,mackeetal2013,rostamietal2016_r2017}.
We assume that our readers are at least vaguely familiar with this kind of model.
In brief, it builds a probability distribution for neuronal activity using
only data about simultaneous firings of two neurons, or of three, and so
on. Let us take the case of the first two moments as an example. The
maximum-entropy distribution can be derived in three main ways:
\begin{enumerate}
\item\label{item:maxent}From the maximum-entropy method
  \cite{jaynes1957,jaynes1963,sivia1996_r2006}. The distribution is chosen
  as the one with the highest Shannon entropy among those having particular
  first and second moments. This distribution is deemed to be
  \enquote{maximally noncommittal} \cite{jaynes1963}, but this adjective
  means little without further technical characterization. Two
  characterizations are given by the next two derivations.

\item\label{item:entropic_prior}As the asymptotic distribution of a Bayesian model called
  \enquote{exchangeable}, with a particular prior
  \cites{jaynes1986d_r1996,skilling1989b,skilling1990,rodriguez1989,rodriguez2002,portamana2017}[see
  also:][\chap~VI]{schroedinger1946_r1948}%[pp.~52--55]
  {jaynes1979b}. This distribution is the limit, when the number of time
  bins is very large, of
  \begin{equation}
    \label{eq:predictive_exchangeability}
    \p(\text{\small activity in one time bin} \|
    \text{\small 1st \amp\ 2nd empirical moments from other bins}%,\text{\small other info}
    ),
  \end{equation}
  calculated with Bayes's theorem from a particular initial joint
  distribution in which different time bins are almost independent. This
  derivation produces a probability just by taking into account those two
  moments, but does not exclude that other information might be relevant.
  It can therefore be compared with a probability that uses additional
  information, or with empirical frequencies.

\item\label{item:sufficiency_model}As the asymptotic distribution of a Bayesian model based on
  sufficient statistics \cite{jaynes1982,%[\sect~I]
    jaynes1982b,portamana2017}. This distribution is the limit, when the
  number of time bins is very large, of
\begin{equation}
  \label{eq:general_prediction}
  \p(\text{\small activity in one time bin} \|
  \text{\small activities in other bins}%,\text{\small other info}
  )
\end{equation}
under the assumption that \emph{only} the values of the first and second
empirical moments of the activities are relevant for the probability above,
\ie\
\begin{multline}
  \label{eq:informal_sufficiency}
  \p(\text{\small activity in one time bin} \|
  \text{\small activities in other bins}%,\text{\small other info}
  ) \equiv{}\\
  \p(\text{\small activity in one time bin} \|
  \text{\small 1st \amp\ 2nd empirical moments from other bins}%,\text{\small other info}
  );
\end{multline}
in other words, the assumption that the first and second moments are
sufficient statistics. This limit generally does not depend on the initial
joint distribution. This derivation produces a probability by explicitly
assuming that only the first and second moments are informationally
relevant: any other information would lead to the same conditional
probability. It can therefore only be compared with empirical frequencies.
\end{enumerate}
We refer our readers to the cited references for a deeper discussion of
these derivations. It is remarkable that all three lead to the same
distribution. We shall now see, however, that if we apply these derivations
more carefully they lead to different distributions.

@@

This kind of investigation of recorded samples of neuronal activity would
have little significance if we did not expect its results to generalize
beyond the particular recorded sample. In other words, we expect that the
conclusions drawn with \me\ methods about the recorded neurons should
somehow generalize or extrapolate at least to larger areas from which those
neurons are recorded. We are thus assuming that the recorded sample is a
\emph{representative sample} of some larger area. Whether such
extrapolation makes biological sense is a difficult question, to be
analysed case by case. For the moment let's assume that it does make
biological sense, and that our sample can be assumed to be representative
of a larger area.

Then a double dilemma appears in the way the \me\ method is typically
applied in the neuroscientific literature to recorded or artificial
samples. In this note we want to discuss one part of this dilemma and offer
a solution to it.

We can apply the \me\ method to the sample, choosing some set of moments,
and generate a probability distribution for the state of the sample. But if
we consider our sample to be representative of a larger population, it
makes sense to apply the \me\ method to the larger population, using one of
the three derivations above, and generate a probability distribution for
the larger population. The distribution for the sample can then be found by
marginalization. Here's the double dilemma:
\begin{itemize}
\item \emph{the distribution obtained from the application at the sample
    level differs from those obtained from the application at the
    population level};
\item \emph{at the population level, the
    derivations~\ref{item:entropic_prior} and~\ref{item:sufficiency_model}
    lead to different distributions} (the derivations \ref{item:maxent}
  and~\ref{item:entropic_prior} always lead to the same distribution).
\end{itemize}

Either application seems to have some commendable features. However,
\emph{the distributions obtained from these two applications differ}. It
goes without saying that if one is \enquote{noncommittal}, the other must
be \enquote{committal}. Which choice is most meaningful?
